---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# killedbypolice

The goal of killedbypolice is to make readily available the data collected by <http://killedbypolice.net/> for exploration, visualisation, and analysis. 

This is an important data set because the 'US government has no comprehensive 
record of the number of people killed by law enforcement.' ([_The Guardian_, 1 June 2015](https://www.theguardian.com/us-news/ng-interactive/2015/jun/01/about-the-counted)). The killedbypolice project is one of three non-government projects that continuously collect data on polic killings (see [Related work] below).

## Installation

You can install killedbypolice from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("benmarwick/killedbypolice")
```

## How to use

This is a basic example which shows you how to access the data in this package:

```{r example}
# load the library
library(killedbypolice)

# load the data frame, ready to work with 
data("kbp2013_2017")

# inspect the data
str(kbp2013_2017)

head(kbp2013_2017)
```

## Some summaries

Here are some summaries of the data:



## Related work 

We don't know much about who collects the data for <http://killedbypolice.net/>, or what their methods are. But there are several Python scripts for scraping it, for example by [markberger](https://github.com/markberger/police-killings-dataset) and [AceLewis](https://github.com/AceLewis/killedbypolice), and a Ruby script by [atom-morgan](https://github.com/atom-morgan/killed-by-police). We have not tried any of these, or used them here. 

There are several other major projects with similar data:

- https://mappingpoliceviolence.org/ contains data on people killed by police since 2013. They aggregate data from [_Fatal Encounters_](http://www.fatalencounters.org/), KilledbyPolice.net, and the U.S. Police Shootings Database. They include data about whether the deceased was armed or not, and if a vehicle was involved, how it contributed to the death. The site features numerous visualisations exploring the data. 
- [_Fatal Encounters_](http://www.fatalencounters.org/) is a database of all deaths through police interaction in the United States since 1 Jan 2000. It is a public [Google spreadsheet](https://docs.google.com/spreadsheets/d/1dKmaV_JiWcG8XBoRgP8b4e9Eopkpgt7FL7nyspvzAsE/edit#gid=0). It contains variables not inlcuded in killedbypolice, such as street addresses and history of mental illness. Data are tweeted at [\@fatalencounters](https://twitter.com/fatalencounters) and [\@the_decedents](https://twitter.com/the_decedents). They are still updating their data. It has been used by [Snopes.com](https://www.snopes.com/do-police-kill-more-whites-than-black-people/).  
- _The Washington Post_ started compiling data on fatal shootings by U.S. police in 2015. Their data are available in CSV files at their [Washington Post GitHub repository](https://github.com/washingtonpost/data-police-shootings). They are still updating their data, and it includes some variables not collected by the killedbypolice, such as whether or not the deceased was armed, and what city the event occured in. These data have also been used by [Buzzfeed](https://github.com/BuzzFeedNews/2015-12-fatal-police-shootings).    
- _The Guardian_ collected data on on fatal shootings by U.S. police in 2015-2016 for their project [The Counted](https://www.theguardian.com/us-news/ng-interactive/2015/jun/01/the-counted-police-killings-us-database). They are no longer updating this project. These data have been used by [FiveThirtyEight](https://github.com/fivethirtyeight/data/tree/master/police-killings) and [Buzzfeed](https://github.com/BuzzFeedNews/2015-12-fatal-police-shootings) and projects on GitHub, e.g. [flother](https://github.com/flother/thecounted) and [Kaggle](https://www.kaggle.com/the-guardian/the-counted).    
- [VICE News](https://news.vice.com/en_us/article/xwvv3a/shot-by-cops) examined both fatal and nonfatal incidents from 2010 through 2016. The data are on [GitHub](https://github.com/vicenews/shot-by-cops/) and in a [Google Sheet](https://docs.google.com/spreadsheets/d/1CaOQ7FUYsGFCHEqGzA2hlfj69sx3GE9GoJ40OcqI9KY/edit#gid=1271324584)

- [Wikipedia](https://en.wikipedia.org/wiki/List_of_killings_by_law_enforcement_officers_in_the_United_States) has lists of killings by law enforcement offiers in the US, but the numbers are much lower than any of the other sources noted here. 



